A Systematic Literature Review on Extensions to Role Based Access Control

### 2013-04-08 paper/sections/abstract.tex
Committing updates based on comments from reading group.

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's, 
RBAC has become an increasingly popular access control mechanism for applications such as 
web applications, mobile computing and databases. RBAC restricts access to resources 
based on the identity of subjects connected to logical groupings of permissions.  
The introduction of computing into new domains has led to proposals for extensions to the standard
RBAC model; for example, defining additional constraints among roles, adding context or defining new heirarchy relationships.
\\
\\
\noindent
{\bf Aim}: \textit{Researchers looking to expand on the body of knowledge surrounding extensions to RBAC need a starting place to prevent re-invention of the wheel.  
Developers need to find and select an access control mode based on their project requirements. 
The goal of this work is to provide practioners an assessment of the current extensions to RBAC, and researchers with insights into the state of RBAC extension models and how they are evaluated.}
\\
\\
\noindent
{\bf Method}: The authors conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collected 1716 papers from electronic digital libraries
IEEE, Google Scholar, CiteSeerX and ACM and selected 29 primary sources.  We performed a comparative 
analysis of the primary sources to find relationships among extended models and analyze the state of RBAC extension evaluations.
\\
\\
\noindent
{\bf Results}: Our results showed that extensions to RBAC fall into a number of categorizations that in turn all
fall under the single category of context based extensions.  We look into the motivations behind RBAC extensions and
the domains that are leading factors in developing these newer models. We also quantify the current evaluation methods
used when RBAC extension models are presented to the research community.
\\
\\
\noindent
{\bf Conclusions}:

>>>>

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's by the National Institute of Standards and Technology (NIST),
computing in new domains has led to proposals for extensions to the standard RBAC model; for example, defining additional constraints among roles, adding context or defining new hierarchy relationships.
\\
\\
\noindent
{\bf Aim}: The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC standard model, and in understanding
how extensions to RBAC are evaluated by providing an assessment of the state of extensions to the RBAC standard model through: establishment of a 
set of extension categories, an examination of the state of the art in evaluations of extension models and a breakdown of the reasoning and motivations
that have led to extension of the RBAC standard.
\\
\\
\noindent
{\bf Method}: We performed a systematic literature review that yielded 1716 papers, of which 28 were deemed primary sources for inclusion as model extensions to the RBAC standard model. 
\\
\\
\noindent
{\bf Results}: Our results show that extensions to RBAC can be classified under eight categories: Constraint, Context, 
Organization, Privacy, Task, Spatio-Temporal, Spatial and Temporal. We found that 12 models presented no evidence
of an evaluation, eight models used example scenarios and 6 others provided complexity and time analysis.  We identified only
eight of the 28 papers provided an implementation of their model in the form of an enterprise application or prototype. We found that the primary domains 
that inspired extensions were the medical domain with nine of the 28 models, enterprise workflows with five of 28, and mobile computing with five of 28.
\\
\\
\noindent
{\bf Conclusions}: The extensions to RBAC found by our systematic literature review fall into eight categories that upon closer inspection can be 
aligned under the single category of context. Our literature review showed that the state of evaluation of RBAC model evaluation is limited with one
paper doing an evaluation comparing itself to the core RBAC model, eight presenting example 
scenarios of their model in action, and 12 of the 28 models lacking any identifiable evaluation.

### 2013-04-08 paper/sections/analysis.tex
Committing updates based on comments from reading group.

Of the 29 papers surveyed, there was a lack of implementation with 21 of the paper providing no
mention of a implementation or prototype.  Of the remaining 8 papers that did mention an implementation, half 
were simply prototypes developed by the authors while the other half were claimed to be implemented within a real
system.

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item What domains or scenarios serve as inspiration for these extensions?
\item Medical domain \cite{alam06:constraint}, \cite{tzelepi01:flexible}, \cite{motta03:contextual}, \cite{ni2010privacy}, \cite{damiani2007geo}, \cite{hansen2003spatial}, \cite{samuel07:spatio-temporal}, \cite{aich09:role}, \cite{zhou2007team}
\item Pervasive computing environments \cite{huang06:pervasive}, \cite{chen08:spatio-temporal}, \cite{ray07:spatio}
\item Web applications \cite{masoumzadeh2008purbac}
\item Mobile computing \cite{thein2011leveraging}, \cite{zou2009crbac}, \cite{chandran05:llt}, \cite{ray07:spatio}, \cite{aich09:role}
\item Large-scale organizations with many sub-departments \cite{yamazaki04:designing}, \cite{han08:extended}, \cite{yao2008task}
\item Enterprise, organization workflows \cite{cholewka00:acontext-sensitive}, \cite{bao08:role}, \cite{zhang06:collaborative}, \cite{oh2003task}, \cite{joshi05:generalized}
\end{itemize}

The predominant domain for which extension models have been generated for is that of the medical domain with 9 of 29 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by 5 papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with 3 each and web applications with 1.  There were 4 papers without any direct mention of a domain
since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

\subsection{Research Question 6}

What commonalities or generalizations exist across all categorizations? \\

>>>>

\begin{table}
\centering
\begin{tabular}{ | c | c | }
\cline{1-2}

\textbf{Implementation Type} & \textbf{Paper Count} \\ \cline{1-2}
Enterprise Implementation & 4 \\ \cline{1-2}
Prototype Implementation & 4 \\ \cline{1-2}
No Implementation & 20 \\

\cline{1-2}
\end{tabular}
\caption{Implementation types found and the count of primary sources}
\label{tab:implementations}
\end{table}

Table \ref{tab:implementations} shows the breakdown of implementations found within the primary sources.
Of the 28 papers surveyed, there was a lack of implementation with 20 of the paper providing no
mention of a implementation or prototype.  Of the remaining 8 papers that did mention an implementation, half 
were simply prototypes developed by the authors while the other half were claimed to be implemented within a real
system.

The RBAC standard was designed with enterprises in mind such that when practioners implemented RBAC into their systems
there would be a reasonable assurance being based off a well thought out model.  As extensions to the standard model
come along, thought and time should be given to how features and nuances of their models may impact implementation
in order to achieve the same goals as the original standard.  The primary sources should a significant lack of implementation
with over 70\% of the models having no notion of attempting to implement them.  The bare minimum, as 4 papers did, should be
a prototype implementation of the model for review by both pracitioners and researchers. Of the models that produced an 
implementation within the enterprise world, two were from within the medical domain and two were implemented using web application technologies.

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Medical domain \cite{alam06:constraint}, \cite{tzelepi01:flexible}, \cite{motta03:contextual}, \cite{ni2010privacy}, \cite{damiani2007geo}, \cite{hansen2003spatial}, \cite{samuel07:spatio-temporal}, \cite{aich09:role}, \cite{zhou2007team}
\item Pervasive computing environments \cite{huang06:pervasive}, \cite{chen08:spatio-temporal}, \cite{ray07:spatio}
\item Web applications \cite{masoumzadeh2008purbac}
\item Mobile computing \cite{thein2011leveraging}, \cite{zou2009crbac}, \cite{chandran05:llt}, \cite{ray07:spatio}, \cite{aich09:role}
\item Large-scale organizations with many sub-departments \cite{yamazaki04:designing}, \cite{han08:extended}, \cite{yao2008task}
\item Enterprise, organization workflows \cite{cholewka00:acontext-sensitive}, \cite{bao08:role}, \cite{zhang06:collaborative}, \cite{oh2003task}, \cite{joshi05:generalized}
\end{itemize}

The predominant domain for which extension models have been generated for is that of the medical domain with 9 of 28 mentioning scenarios or requirements of that industry.
Mobile computing and enterprise workflows were each represented by 5 papers claiming to be influenced by the requirements for access control within these domains. The final set
of domains was pervasive computing environments and large-scale organizations with 3 each and web applications with 1.  There were 4 papers without any direct mention of a domain
since Aich et al. \cite{aich09:role} fall under both the medical domain and mobile computing.

@@

RQ6: What commonalities or generalizations exist across all categorizations? \\

### 2013-04-08 paper/sections/core_rbac.tex
Adding an updated introduction.
Committing updates based on comments from reading group.

Since the basis for our review is extensions to the core model, we will describe the core model, associated entities and other terminology encountered across the space of our review.  The NIST RBAC model proposed by Ferraiolo et al. \cite{ferraiolokuhn} and later adopted as the official standard for RBAC by the International Committee for Information Technology Standards (INCITS) consists of four basic entities:

@@

In RBAC, a user can exercise a permission only if the user is assigned to a role.
In addition to the four basic components, two functions are defined:
the user role assignment (UA) and the role
permission assignment (PA) functions.
UA models assignment of users to roles.
PA models assignment of permissions to roles.

@@

>>>>

RBAC provides effective and efficient permissions management for operations, especially when sharing resources in an organizations.
Prior to the creation of the NIST RBAC standard, no general agreement on the definition 
of RBAC existed amongst practioners or within the research community. 
Without a unified definition of RBAC, software developers describe similar concepts and features of RBAC models using different terminology, which was shown to slow implementation of RBAC.  
Moreover, in cases where organizations are concerned with adopting RBAC,
evaluation and comparison of RBAC technologies developed by different vendors was difficult.
NIST, in collaboration with industry and academics, worked on defining a set of consensus RBAC concepts and terminology and proposed a standard for RBAC that addressed these cost and interoperability issues by developing a common definition that can be used across different vendors.

NIST's work can directly benefit organizations by lowering cost of early phase R\&D and implementation costs of RBAC.
Since RBAC is introduced, 2010 RTI report showed that the the rate of RBAC adoption is rapidly growing for recent years. 
The analysts estimate that, by the end of 2010, at least a part of permission of systems use RBAC for more than 50\% of IT users at organizations, which
hire more than 500 employees. For economic estimate analysis, the analysts estimate that RBAC technology has generated \$6.1 billion in net economic benefits to industry, of which \$1.1 billion is attributable to NIST RBAC standard work.

Since the basis for our review is extensions to the core model, we will describe the core model, associated entities and other terminology encountered across the space of our review.

\subsection{Core RBAC}

The NIST RBAC model proposed by Ferraiolo et al. \cite{ferraiolo} and later adopted as the official standard for RBAC by the INCITS includes three types of RBAC models: core RBAC, hierarchal RBAC, and constrained RBAC. The four entities of the standard RBAC model are:

@@

In RBAC, a user can exercise a permission only if the user is assigned to a role.
In addition to the four basic components, two functions are defined:
the user role assignment (UA) and the role and permission assignment (PA) functions.
User role assignment models assignment of users to roles.
Permission role assignment models assignment of permissions to roles.

@@

\subsection{Hierarchal RBAC}

Hierarchal RBAC model adds role hierarchies (RH) feature to core RBAC model.
This model incorporates a structure of roles in an organization using inheritance relationship among attributes such as roles.
The role structure in an organization may use
a role $r_1$, who inherits all permissions of another role $r_2$.
For example, a manager role may inherit all permissions of a employee role.
Role hierarchies help simplify access control policy creation and maintenance by reducing the number of
individual role assignments to a user. Formally, role inheritance relation is shown as \textit{RH} $\subseteq$ \textit{ROLES} $\times$ \textit{ROLES} describing many-to-many mapping role inheritance relation. 
General role hierarchies can be extended to use the concept of multiple inheritance where
a role $r_1$, who inherits all permissions from more than one roles.

\subsection{Constrained RBAC}

Constrained RBAC incorporates static and dynamic separation of duty relations to the RBAC model. Separation of 
duty relations enforce conflict of interest among roles. This model defines two types of separation of duty relations; static and dynamic.

\begin{itemize}
	\item Static Separation of Duty (SSoD): SSoD restricts the conflicting-role assignments statically that are associated with a user. On situations
	where multiple roles can be associated with a single user and roles $Role_A$ and $Role_B$ are conflicting each other, no permission is given to a user who is assigned to both $Role_A$ and $Role_B$. SSoD is known to be too rigid for practical use in cases where a user should have permissions as either $Role_A$ and $Role_B$.
	\item Dynamic Separation of Duty (DSD): Dynamic SoD is known to be
more flexible than SSD. DSD restricts the conflicting-role assignments dynamically that are associated with a user. On situations
	where multiple roles can be associated with a single user and , given a context, roles $Role_A$ and $Role_B$ are conflicting each other dynamically, no permission is given to a user.	
\end{itemize}
